Live Functional Programming with Typed Holes

### 2018-10-31 calculus.tex
small clarifications through sec 3, addressing R2's confusion about matching relation, and more editor states being meaningful; also changed the description of the mechanization to be consistent with our final artifact

\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style
%
with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We start with synthesis for the programmer's ``top level'' external
expression.

@@

Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$.
For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.
%
To avoid the need for two separate rules, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.}
%
% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that
%%

The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matched arrow type judgement.

@@

The cast form in Hazelnut Live is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
%
This form serves to ``box'' an expression of type $\htau_1$ for
treatment as an expression of a consistent type $\htau_2$
(Rule~\rulename{TACast} in \Figref{fig:hasType}).%
\footnote{
In the earliest work on gradual type theory, the cast form only gave
the target type~$\htau_2$ \cite{Siek06a}, but it simplifies the dynamic semantics substantially
to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
}

Elaboration inserts casts at function applications and ascriptions.
%
The latter is more straightforward: Rule~\rulename{ESAsc}
in \Figref{fig:elabSyn} inserts a cast from the assigned type to the
ascribed type.
%
Theorem~\ref{thm:typed-elaboration} inductively ensures that the two types
are consistent.
%
We include ascription for exposition purposes---this form is derivable
by using application together with the half-annotated identity, $e
: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
elaboration, discussed below, is more general.

Rule~\rulename{ESAp} elaborates function applications.
%
To understand the rule, consider the elaboration of external
expression~$\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$, the example
discussed above:
\[
        \hap{\dcasttwo{
        (\halam{x}{\tehole}{\underbrace{
                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
                {\dcasttwo{c}{b}{\tehole}}
                }_{\textrm{elaboration of function body}~x(c)}
        }
        )}{\tarr{\tehole}{\tehole}}{
           \tarr{\tehole}{\tehole}}
           }
           {\dcasttwo{c}{b}{\tehole}}
\]
Consider the (indicated) function body,
%
where elaboration inserts a cast on both the function expression~$x$ and
its argument~$c$.
%
Together, these casts for~$x$ and~$c$ permit assigning a type to the
function body according to the rules in \Figref{fig:hasType}, where we
could not do so under the same context without casts.
%
We separately consider the elaborations of~$x$ and of~$c$.

@@

Next, consider the application's argument, here constant~$c$.
%
The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
elaboration, from the type it is assigned by the final premise of the
rule~(type~$b$ here), to the argument type of the matched arrow type of the
function expression~(type~$\tehole$ here).

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName for the full definition).
% Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} rules just described only operate at ground type.
%
The two remaining instruction transition rules, Rule~\rulename{ITGround}
and~\rulename{ITExpand}, insert intermediate casts from non-ground type to a
consistent ground type, and \emph{vice versa}.
%
These rules serve as technical devices, permitting us to restrict our
interest exclusively to casts involving ground types and type holes elsewhere.
%
Here, the only non-ground types are the arrow types, so the grounding
judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
%
More generally, the following invariant governs this judgement.
\begin{lem}[Grounding]
  If $\groundmatch{\htau_1}{\htau_2}$
  then $\isGround{\htau_2}$
  and $\tconsistent{\htau_1}{\htau_2}$
  and $\htau_1\neq\htau_2$.
\end{lem}

In all other cases, casts evaluate either to boxed values or to
indeterminate forms according to the remaining rules
in \Figref{fig:isFinal}.
%
Of note, Rule \rulename{ICastHoleGround} handles casts from hole to
ground type that are not of
the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$.
%
The need for this case is perhaps not obvious because in the cast calculus,  the only irreducible term of type $\tehole$
is $\dcasttwo{\dexp}{\htau_1}{\tehole}$, but here there are holes as well.

@@

\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}
\noindent
%
The proof relies on an analogous preservation lemma for instruction
transitions and a standard substitution lemma stated in
the \appendixName.
% Appendix~\ref{sec:additional-defns}.
%
Hole closures can disappear during evaluation, so we rely on simple weakening
of $\Delta$.

@@

The supplemental material includes our Agda
mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk}
of the semantics and metatheory of \HazelnutLive,
%including proofs of all of the
including all of the theorems stated above and necessary lemmas.
We take as assumptions only a few standard properties of general hypothetical judgements, substitution and finite maps.
%
%We choose the
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common).
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}).
%
%
Our approach is otherwise standard: we model judgements as
inductive datatypes, and rules as dependently typed constructors of these judgements.
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names and encode typing
contexts, hole contexts and finite substitutions using metafunctions.
To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}.
The documentation provided with the mechanization has more details.

@@

The \Hazel implementation demonstrated in Sec.~\ref{sec:examples}
includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
in this section, albeit for an \Elm-like language that goes somewhat beyond the simple
calculus here. As with many full-scale systems, there is not currently a formal
specification for the full \Elm{} or \Hazel language, but the \appendixName
discusses how the standard approach for deriving a ``gradualized'' version of a
language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples.

@@

The editor component of the \Hazel implementation is derived
from the structure editor calculus of~\Hazelnut, but with support for more natural cursor-based movement and infix operator sequences (the details of which are beyond the scope of this paper). It exposes a language of structured
edit actions that automatically insert empty and non-empty holes as necessary
to guarantee that every edit state has some (possibly incomplete) type. This corresponds to the top-level Sensibility invariant established for the \Hazelnut calculus by \citet{popl-paper}, reproduced below:
\begin{prop}[Sensibility]
  \label{thrm:actsafe}
  If $\hsyn{\Gamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\Gamma}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{prop}
\noindent
Here, $\zexp$ is an editor state (an expression with a cursor), and $\removeSel{\zexp}$ drops the cursor, producing an expression ($e$ in this paper). So in words, ``if, ignoring the cursor, the editor state, $\removeSel{\zexp}$, initially has type $\htau$ and we perform an edit action $\alpha$ on it, then the resulting editor state, $\removeSel{\zexp'}$, will have type $\htau'$''.

By composing this Sensibility property with the \Property{Elaborability},
\Property{Typed Elaboration}, \Property{Progress} and
\Property{Preservation} properties from this section, we establish a
uniquely powerful Continuity invariant:
\begin{corol}[Continuity]
  \label{thrm:actsafe}
  If $\hsyn{\emptyset}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
    $\elabSyn{\emptyset}{\removeSel{\zexp'}}{\htau'}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$ such that
$\hasType{\Delta}{\emptyset}{\dexp}{\htau'}$
and either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ for some $d'$ such that $\hasType{\Delta}{\emptyset}{\dexp'}{\htau'}$; or
  (b) $\isBoxedValue{\dexp}$ or
  (c) $\isIndet{\dexp}$.
  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{corol}

@@

In settings where the editor does not maintain this Sensibility invariant, our approach still helps to reduce the severity of the gap problem, i.e. \emph{more} editor states are meaningful, even if not \emph{all} are.% We can formally state this end-to-end continuity corollary as follows:
%
% We make no further claims about the usability or practicality of the implementation (and indeed, there remain many important and decidedly unresolved questions along these lines, which we leave beyond the scope of this paper).
% We also reiterate that the essential ideas developed from type-theoretic first principles in this paper do not require that
% the editor component of the programming environment be implemented as a structure editor (see Sec.~\ref{sec:intro}).
 %,  to scale up these ideas to a ``real-world'' language and programming environment.

>>>>

\Figref{fig:bidirectional-typing} defines the type system in the \emph{bidirectional} style
%
with two mutually defined judgements \cite{Pierce:2000ve,bidi-tutorial,DBLP:conf/icfp/DunfieldK13,Chlipala:2005da}. The type synthesis
judgement~$\hsyn{\hGamma}{\hexp}{\htau}$ synthesizes a type~$\htau$
for external expression~$\hexp$ under typing context $\hGamma$, which tracks typing
assumptions of the form $x : \htau$ in the usual
manner \cite{pfpl,tapl}.
%
The type analysis judgement~$\hana{\hGamma}{\hexp}{\htau}$ checks
expression~$\hexp$ against a given type~$\htau$.
%
Algorithmically, analysis accepts a type as input, and synthesis gives
a type as output.
%
We start with synthesis for the programmer's top level external
expression.

@@

Rule \rulename{SVar} synthesizes the corresponding type from $\hGamma$.
For the sake of exposition, \HazelnutLive includes ``half-annotated'' lambdas, $\halam{x}{\htau}{\hexp}$, in addition to the unannotated lambdas, $\hlam{x}{\hexp}$, from \Hazelnut.
%
Half-annotated lambdas may appear in synthetic position according to Rule \rulename{SLam}, which is standard \cite{Chlipala:2005da}.
%
Unannotated lambdas may only appear where the expected type is known to be either an arrow type or the hole type, which is treated as if it were $\tarr{\tehole}{\tehole}$.\footnote{A system supporting ML-style type reconstruction \cite{damas1982principal} might also include a synthetic rule for unannotated lambdas, e.g. as outlined by \citet{DBLP:conf/icfp/DunfieldK13}, but we stick to this simpler ``Scala-style'' local type inference scheme in this paper \cite{Pierce:2000ve,Odersky:2001lb}.} 
%
To avoid the need for separate rules for these two cases, Rule \rulename{ALam} uses the matching relation $\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}$ defined in \Figref{fig:arrmatch}, which produces the matched arrow type $\tarr{\tehole}{\tehole}$ given the hole type, and operates as the identity on arrow types \cite{DBLP:conf/snapl/SiekVCB15,DBLP:conf/popl/GarciaC15}.
%
% \Secref{sec:related} dicusses how \HazelnutLive might be enriched with
% with ML-style type reconstruction~\cite{damas1982principal}, perhaps via
% the approach outlined by~\citet{DBLP:conf/icfp/DunfieldK13}.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% To Cyrus from Matt:
%
% Why say the following here? --- It's discussing a different design that we didn't pursue here.  We should move such discussion to related work.
%
%Note that
%%

The rule governing function application, Rule \rulename{SAp}, similarly treats an expression of hole type in function position as if it were of type $\tarr{\tehole}{\tehole}$ using the same matching relation.

@@

The cast form in \HazelnutLive is $\dcasttwo{\dexp}{\htau_1}{\htau_2}$.
%
This form serves to ``box'' an expression of type $\htau_1$ for
treatment as an expression of a consistent type $\htau_2$
(Rule~\rulename{TACast} in \Figref{fig:hasType}).%
\footnote{
In the earliest work on gradual type theory, the cast form only gave
the target type~$\htau_2$ \cite{Siek06a}, but it simplifies the dynamic semantics substantially
to include the assigned type~$\htau_1$ in the syntax \cite{DBLP:conf/snapl/SiekVCB15}.
}

Elaboration inserts casts at function applications and ascriptions.
%
The latter is more straightforward: Rule~\rulename{ESAsc}
in \Figref{fig:expandSyn} inserts a cast from the assigned type to the
ascribed type.
%
Theorem~\ref{thm:typed-elaboration} inductively ensures that the two types
are consistent.
%
We include ascription for expository purposes---this form is derivable
by using application together with the half-annotated identity, $e
: \tau = \hap{(\halam{x}{\htau}{x})}{e}$; as such, application
elaboration, discussed below, is more general.

Rule~\rulename{ESAp} elaborates function applications.
%
To understand the rule, consider the elaboration of the example
discussed above, $\hap{(\halam{x}{\tehole}{\hap{x}{c}})}{c}$:
\[
        \hap{\dcasttwo{
        (\halam{x}{\tehole}{\underbrace{
                \hap{\dcasttwo{x}{\tehole}{\tarr{\tehole}{\tehole}}}
                {\dcasttwo{c}{b}{\tehole}}
                }_{\textrm{elaboration of function body}~x(c)}
        }
        )}{\tarr{\tehole}{\tehole}}{
           \tarr{\tehole}{\tehole}}
           }
           {\dcasttwo{c}{b}{\tehole}}
\]
Consider the (indicated) function body,
%
where elaboration inserts a cast on both the function expression~$x$ and
its argument~$c$.
%
Together, these casts for~$x$ and~$c$ permit assigning a type to the
function body according to the rules in \Figref{fig:hasType}, where we
could not do so under the same context without casts.
%
We separately consider the elaborations of~$x$ and of~$c$.

@@

Next, consider the application's argument, here constant~$c$.
%
The conclusion of Rule~\rulename{ESAp} inserts the cast on the argument's
elaboration, from the type it is assigned by the final premise of the
rule~(type~$b$), to the argument type of the matched arrow type of the
function expression~(type~$\tehole$).

@@

Substitution, written $[d/x]d'$, operates in the standard capture-avoiding manner~\cite{pfpl} (see the \appendixName~for the full definition).
% Appendix~\ref{sec:additional-defns}).
%
The only cases of special interest arise when substitution reaches a hole closure:
\[
\begin{array}{rcl}
  [d/x]\dehole{u}{\sigma}{} & = & \dehole{u}{[d/x]\sigma}{} \\%
  \substitute{d}{x}{\dhole{d'}{u}{\sigma}{}} & = & \dhole{[d/x]d'}{u}{[d/x]\sigma}{}
\end{array}
\]
In both cases, we write~$[d/x]\sigma$ to perform substitution on each expression in the hole environment~$\sigma$, i.e. the environment ``records'' the substitution.
%
For example, $\stepsToD{}
    {\hap{(\halam{x}{b}{\halam{y}{b}{\dehole{u}{[x/x, y/y]}{}}})}{c}}
    {\halam{y}{b}{\dehole{u}{[c/x, y/y]}{}}}$.
%
Beta reduction can duplicate hole closures.
%
Consequently, the environments of different closures with the same hole name may differ,
e.g., when a reduction applies a function with a hole closure body multiple times as in Fig.~\ref{fig:grades-example}.
Hole closures may also appear within the environments of other hole
closures, giving rise to the closure paths described in
Sec.~\ref{sec:paths}.

@@

% Rules \rulename{ITCastSucceed} and \rulename{ITCastFail} only operate at ground type.
%
The two remaining instruction transition rules, Rule~\rulename{ITGround}
and~\rulename{ITExpand}, insert intermediate casts from non-ground type to a
consistent ground type, and \emph{vice versa}.
%
These rules serve as technical devices, permitting us to restrict our
interest exclusively to casts involving ground types and type holes elsewhere.
%
Here, the only non-ground types are the arrow types, so the grounding
judgement~$\groundmatch{\tau_1}{\htau_2}$ (\Figref{fig:groundmatch}),
produces the ground arrow type~$\tarr{\tehole}{\tehole}$.
%
More generally, the following invariant governs this judgement.
\begin{lem}[Grounding]
  If $\groundmatch{\htau_1}{\htau_2}$
  then $\isGround{\htau_2}$
  and $\tconsistent{\htau_1}{\htau_2}$
  and $\htau_1\neq\htau_2$.
\end{lem}

In all other cases, casts evaluate either to boxed values or to
indeterminate forms according to the remaining rules
in \Figref{fig:isFinal}.
%
Of note, Rule \rulename{ICastHoleGround} handles casts from hole to
ground type that are not of
the form $\dcastthree{\dexp}{\htau_1}{\tehole}{\htau_2}$.
%

@@

\begin{thm}[Preservation]
  If $\hasType{\Delta}{\emptyset}{\dexp}{\htau}$ and
  $\stepsToD{\Delta}{\dexp}{\dexp'}$ then
  $\hasType{\Delta}{\emptyset}{\dexp'}{\htau}$.
\end{thm}
\noindent
%
The proof relies on an analogous preservation lemma for instruction
transitions and a standard substitution lemma stated in
the \appendixName.
% Appendix~\ref{sec:additional-defns}.
%
Hole closures can disappear during evaluation, so we must have structural weakening 
of $\Delta$.

@@

The supplemental material includes our Agda
mechanization  \cite{norell2009dependently,norell:thesis,Aydemir:2005fk}
of the semantics and metatheory of \HazelnutLive,
%including proofs of all of the
including all of the theorems stated above and necessary lemmas.
%
%We choose the
%(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common).
%
%Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation.
%
%Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}).
%
%
Our approach is standard: we model judgements as
inductive datatypes, and rules as dependently typed constructors of these judgements.
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and hole names, and avoid certain other complications relted to substitution by enforcing the requirement that all bound variables in a term are unique when convenient (this requirement can always be discharged by alpha-variation). We encode typing
contexts and hole contexts using metafunctions.
To support this encoding choice, we postulate function extensionality (which is independent of Agda's axioms) \cite{awodey2012inductive}. We encode finite substitutions as an inductive datatype with a base case representing the identity substitution and an inductive case that records a single substitution. Every finite substitution can be represented in this scheme. This makes it easier for Agda to see why certain inductions that we perform are well-founded.  
The documentation provided with the mechanization has more details.

@@

The \Hazel implementation described in Sec.~\ref{sec:examples}
includes an unoptimized interpreter, written in OCaml, that implements the semantics as described
in this section, with some simple extensions. As with many full-scale systems, there is not currently a formal
specification for the full \Hazel language, but the \appendixName
discusses how the standard approach for deriving a ``gradualized'' version of a
language construct provides most of the necessary scaffolding \cite{DBLP:conf/popl/CiminiS16}, and provides some examples (sum types and numbers).

@@

The editor component of the \Hazel implementation is derived
from the structure editor calculus of~\Hazelnut, but with support for more natural cursor-based movement and infix operator sequences (the details of which are beyond the scope of this paper). It exposes a language of structured
edit actions that automatically insert empty and non-empty holes as necessary
to guarantee that every edit state has some (possibly incomplete) type. This corresponds to the top-level Sensibility invariant established for the \Hazelnut calculus by \citet{popl-paper}, reproduced below:
\begin{prop}[Sensibility]
  \label{thrm:sensibility}
  If $\hsyn{\Gamma}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\Gamma}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
    $\hsyn{\hGamma}{\removeSel{\zexp'}}{\htau'}$.
  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{prop}
\noindent
Here, $\zexp$ is an editor state (an expression with a cursor), and $\removeSel{\zexp}$ drops the cursor, producing an expression ($e$ in this paper). So in words, ``if, ignoring the cursor, the editor state, $\removeSel{\zexp}$, initially has type $\htau$ and we perform an edit action $\alpha$ on it, then the resulting editor state, $\removeSel{\zexp'}$, will have type $\htau'$''.

By composing this Sensibility property with the \Property{Elaborability},
\Property{Typed Elaboration}, \Property{Progress} and
\Property{Preservation} properties from this section, we establish a
uniquely powerful Continuity invariant:
\begin{corol}[Continuity]
  \label{thrm:continuity}
  If $\hsyn{\emptyset}{\removeSel{\zexp}}{\htau}$ and
    $\performSyn{\emptyset}{\zexp}{\htau}{\alpha}{\zexp'}{\tau'}$ then
    $\elabSyn{\emptyset}{\removeSel{\zexp'}}{\htau'}{\dexp}{\Delta}$
      for some $\dexp$ and $\Delta$ such that
$\hasType{\Delta}{\emptyset}{\dexp}{\htau'}$
and either
  (a) $\stepsToD{}{\dexp}{\dexp'}$ for some $d'$ such that $\hasType{\Delta}{\emptyset}{\dexp'}{\htau'}$; or
  (b) $\isBoxedValue{\dexp}$ or
  (c) $\isIndet{\dexp}$.
  % \item If $\hana{\hGamma}{\removeSel{\zexp}}{\htau}$ and
  %   $\performAna{\hGamma}{\zexp}{\htau}{\alpha}{\zexp'}$ then
  %   $\hana{\hGamma}{\removeSel{\zexp'}}{\htau}$.
\end{corol}

@@

In settings where the editor does not maintain this Sensibility invariant, but where programmers can manually insert holes, our approach still helps to reduce the severity of the gap problem, i.e. \emph{more} editor states are dynamically meaningful, even if not \emph{all} of them are.% We can formally state this end-to-end continuity corollary as follows:
%
% We make no further claims about the usability or practicality of the implementation (and indeed, there remain many important and decidedly unresolved questions along these lines, which we leave beyond the scope of this paper).
% We also reiterate that the essential ideas developed from type-theoretic first principles in this paper do not require that
% the editor component of the programming environment be implemented as a structure editor (see Sec.~\ref{sec:intro}).
 %,  to scale up these ideas to a ``real-world'' language and programming environment.
